---
title: "HW 01"
author: "Cat Xia"
format: 
  html:
    embed-resources: true
toc: true
---

## 0 - Setup

```{r setup}
if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(tidyverse, here, glue, scales, openintro) 

devtools::install_github("tidyverse/dsbox")
```

## 1 - Road traffic accidents in Edinburgh
```{r Load Data: accidents}
accidents_file_loc <- here("data", "accidents.csv")
accidents <- read.csv(accidents_file_loc)
#accidents |> 
  #glimpse()
```

```{r Data Wrangling: accidents}
#create column that separates weekdays from weekend. Change time to the hour minute second format
accidents <- accidents |>
  mutate(
    day_of_week_cat = ifelse(day_of_week %in% c("Saturday", "Sunday"), "Weekend", "Weekday"),
    time = hms(time),
    time_sec = as.numeric(time)
  )
    #code for ifelse from https://stackoverflow.com/questions/28893193/creating-factor-variables-weekend-and-weekday-from-date
```


```{r Accidents graph}
accidents |>
  #set x as time in the day, and group as severity
  ggplot(aes(x = time_sec, group = severity, fill = severity)) +
  
  #create a density plot with 50% fill
  geom_density(alpha = 0.7) +
  
  #set manual colors for the graph
  scale_fill_manual(values = c("Fatal" = "#aa93b0", "Serious" = "#9ecac8", "Slight" = "#fef39f")) +
  #Hexcodes taken from online color picker from the original picture
  
  #separate graphs based on weekend or weekday
  facet_wrap(~day_of_week_cat, nrow = 2) +
  
  #labels
  labs(
    x = "Time of day",
    y = "Density",
    fill = "Severity",
    title = "Number of accidents throughout the day",
    subtitle = "By day of week and severity",
    caption = "Source: Data.gov.uk (2018)"
  ) +
  
  scale_x_continuous(breaks = c(0, 14400, 28800, 43200, 57600, 72000, 86400),
                     
    # convert seconds back to HH:MM:SS                
    labels = function(x) {
    hours <- x %/% 3600
    minutes <- (x %% 3600) %/% 60
    seconds <- x %% 60
    sprintf("%02d:%02d:%02d", hours, minutes, seconds)
    #code for sprintf consulted with AI for formatting issues after attempting with https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sprintf
  })+
  
  theme_minimal()

```

The graph illustrates the density of traffic accidents based on the time of day with respect to weekday or weekends. It evaluates these points based on the level of severity of said accident, with each accident being classified as either Fatal, Serious, or Slight. The data clearly shows that although the amount of Serious and Slight traffic accidents are similar between the weekday and weekend, the number of fatal accidents on the weekdays are significantly greater. Additionally, regardless of the day of week, most car crashes seem to occur between 12PM and 9PM, with a peak at around 4PM.

## 2 - NYC marathon winners

```{r Load Data: Marathon}
#Loads the data for marathon
marathon_file_loc <- here("data", "nyc_marathon.csv")
marathon <- read.csv(marathon_file_loc)
#marathon |> 
  #glimpse()
```

```{r Data Wrangling: Marathon A}
marathon <- marathon |>
  mutate(
    time = hms(time),
  ) |>
  drop_na(time)
```

```{r Graphing: Marathon A}
marathon |>
  ggplot(aes(x = time_hrs)) +
  geom_histogram(bins = 20) +
  labs(
    x = "Time (hours)",
    y = "Number of runners",
    title = "Distribution of Marathon Winner Completion Time",
    subtitle = "Histogram",
    caption = "Source: NYC Marathon | Openintro"
  ) + 
  theme_minimal()

marathon |>
  ggplot(aes(x = time_hrs)) +
  geom_boxplot() + 
  #coord_cartesian(ylim = c(0,30)) +
  labs(
    x = "Time (hours)",
    #y = "Number of runners",
    title = "Distribution of Marathon Winer Completion Time",
    subtitle = "Boxplot",
    caption = "Source: NYC Marathon | Openintro"
  ) +
  theme_minimal()
```
The histogram and boxplot both show that number of runners who completed the marathon after 2.6 hours is very minimal. However, the histogram is able to show that there is a portion of time briefly after 2.25 hours where few to no runners completed the marathon, but the majority of runners completed the marathon at times around then. Additionally, it is able to show the number of runners for each time point. The boxplot, on the other hand, better shows the actual median hours it takes for runners to complete the marathon and the first and third quartile, and thus we can see that more runners finished in between the first quartile and median than the median and third quartile.

```{r}
marathon |>
  ggplot(aes(x = time_hrs, color = division)) +
  geom_boxplot() + 
  facet_wrap(~division) +
  scale_color_manual(values = c("deepskyblue", "salmon4")) +
  labs(
    x = "Time (hours)",
    title = "Distribution of Marathon Winner Completion Time",
    subtitle = "By gender",
    color = "Gender",
    caption = "Source: NYC Marathon | Openintro"
  ) +
  theme_minimal()
```
The median and majority of male runners finish the marathon before minimum time of female runners, with a tighter distance between the outliers than the female runners. Additionally, the more male runners fell between the median and third quartile than fell between the first quartile and median, while female runners were mostly equal. Lastly, both men and women had outliers to the right of the median but none or few to the left.

```{r}
marathon |>
  ggplot(aes(x = time_hrs, color = division)) +
  geom_boxplot() + 
  facet_wrap(~division) +
  coord_cartesian(ylim = c(0, 0.4)) +
  scale_color_manual(values = c("deepskyblue", "salmon4")) +
  labs(
    x = "Time (hours)",
    title = "Distribution of Marathon Winner Completion Time",
    subtitle = "By Gender",
    color = "Gender",
    caption = "Source: NYC Marathon | Openintro"
  ) +
  theme_minimal()
```
The part of the box that goes below 0 is redundant. This increases the data to ink ratio because it gets rid of the extra empty coordinate space below the 0.

```{r}
marathon |>
  ggplot(aes(x = year, y = time_hrs, color = division)) +
  geom_line() + 
  facet_wrap(~division) +
  scale_color_manual(values = c("deepskyblue", "salmon4")) +
  labs(
    x = "Year",
    y = "Time (hours)",
    color = "Gender",
    title = "Marathon winner competion time from 1970 to 2020",
    subtitle = "By gender",
    caption = "Source: NYC Marathon | Openintro"
  ) +
  theme_minimal()
```
The time series shows that not only do both men and women winners show a decrease in marathon time since 1970, but that women winners show a significant decrease in time. The boxplot and histogram can only show the data for all the different years combined.

## 3 - US counties

```{r}

```


## 4 - Rental apartments in SF
```{r Load Data: Credit}
#loads credit data
credit_file_loc <- here("data", "credit.csv")
credit <- read.csv(credit_file_loc)
#credit |> 
  #glimpse()
```


## 5 - Napoleon’s march.

```{r Load Data: Napoleon}
#Loads up napoleon list data
napoleon_file_loc <- here("data", "napoleon.rds")
napoleon <- read_rds(napoleon_file_loc)
#napoleon |>
  #glimpse()
```

